\thetitle

### 2021-06-01 writing/why-rational.tex
ben edits
- delete "see" as in "see sec. 1"; in most places, there's no need for this
  extra direction
- "mix-typed" -> "mixed-typed"
- move ",here ..." clause to directly follow target
- add oxford comma
- plural -> singular b/c pack-lib is the one and only responsible party
- use "--" for Wadler--Findler
- rational-stuff: **A successful** trail
- rational-transient: "and each" -> "and that each" to clarify the assignment
  says two things .... but it's a little strange to say that shallow checking
  is part of what the blame says
- "blame set" -> "blame sequence"
- benchmarks programs were

This paper generalizes Lazarek et. al.'s idea to {\em the rational
programmer\/}.  Like~\citet{mill1874essays}'s {\it homo economicus\/}, the
rational programmer approximates the behavior of a software developer who
reduces time spent on a task by exploiting the available information. In the
context of gradual typing, the rational programmer has two pieces of information
when an impedance mismatch signals exceptional behavior: the error message and
the state of the program. Hence, the most rational procedure is to use the
former to improve the latter. Specifically, the rational programmer translates
the Wadler-Findler slogan into a debugging method, searching for the source of
the impedance mismatch by adding type annotations to some of the untyped parts
of the program identified in the error.  If the type checker rejects an
annotation derived from the context, the rational programmer has found the
source of the problem.  Otherwise, the rational programmer concludes that the
just-annotated parts are not the problem and re-runs the program---which must,
by the slogan, blame a different location for the problem. At this point, the
rational programmer can iterate the process.  Measuring this simulated behavior
on a large number of scenarios yields data that is similar to data collected in
a human-facing study.

@@

Like {\it homo economicus\/}, the rational programmer is an approximation.
People do not behave in a purely rational manner as economic actors, and they
also do not do so as software developers. The point is not to deny the existence
of ``lucky hunches'' programmers or ``tinkering works'' approaches and so on. It
is also not to claim that equipping entire modules with types represents an
always feasible approach\footnote{Adding types at the expression level, say, as
in TypeScript should be considered well within bounds.}  {\em But\/}, the
concept of studying the idea of an economically rational actor has produced
benefits to the discipline of political economics, and this paper suggests that
implementing and studying the rational programmer will help language designers.

>>>>

This paper generalizes Lazarek et. al.'s idea to {\em the rational
programmer\/}.  Like~\citet{mill1874essays}'s {\it homo economicus\/}, the
rational programmer approximates the behavior of a software developer who
reduces time spent on a task by exploiting the available information. In the
context of gradual typing, the rational programmer has two pieces of information
when an impedance mismatch signals exceptional behavior: the error message and
the state of the program. Hence, the most rational procedure is to use the
former to improve the latter. Specifically, the rational programmer translates
the Wadler--Findler slogan into a debugging method, searching for the source of
the impedance mismatch by adding type annotations to some of the untyped parts
of the program identified in the error.  If the type checker rejects an
annotation derived from the context, the rational programmer has found the
source of the problem.  Otherwise, the rational programmer concludes that the
just-annotated parts are not the problem and re-runs the program---which must,
by the slogan, blame a different location for the problem. At this point, the
rational programmer can iterate the process.  Measuring this simulated behavior
on a large number of scenarios yields data that is similar to data collected in
a human-facing study.

@@

Like {\it homo economicus\/}, the rational programmer is an approximation.
People do not behave in a purely rational manner as economic actors, and they
also do not do so as software developers. The point is not to deny the existence
of ``lucky hunches'' programmers or ``tinkering works'' approaches and so on. It
is also not to claim that equipping entire modules with types represents an
always feasible approach.\footnote{Adding types at the expression level, say, as
in TypeScript should be considered well within bounds.}  {\em But\/}, the
concept of studying the idea of an economically rational actor has produced
benefits to the discipline of political economics, and this paper suggests that
implementing and studying the rational programmer will help language designers.
